---
title: "Gridded Line Transect Guide"
author: "Taiki Sakai"
date: "9/21/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Distance sampling techniques we need a way to quantify how much effort went
into finding each detection. Normally with line transect surveys this is done 
in 3 steps. First we estimate a detection function function from all of our detections - this
lets us estimate how far out we can detect animals. Then we break our survey trackline 
into segments, and assign each detection to the segment in which it was detected. 
Then for each segment we compute the approximate area surveyed by multiplying the 
average detection distance from the deteciton function by the length of the trackline.
This method has historically worked well visual line transect surveys, but for
acoustic detections there are some potential problems. When using line transect data
to create spatial density models, it is useful to assign environmental covariates like
sea surface temperature or salinity to each detection. Since each detection is associated
with a segment of trackline, we can use the location of the segment to get the matching
environmental data. For visual surveys this is fine, since visual detection ranges are
small relative to the scale at which environmental conditions may change (e.g. the
sea surface temperature is unlikely to change across the distance that you can spot
a whale). However, acoustics has the ability to detect certain species from upwards
of 15 kilometers away. Across such a large distance, it is definitely possible (and
in some parts of the world, very likely) that environmental variables will change. 
For this reason we want a method of assigning effort to each detection that will also
allow us assign environmental variables that are closer to the location of the call
rather than just on the trackline. 

To solve this issue, we will change the way that we assign effort to detections. Rather
than breaking our study area into trackline segments and then assigning detections to
each segment, we will take our entire study area and break it up in to a grid of smaller
squares. For each square in the grid we will create an estimate of the amount of effort
we have surveyed that square using the same detection function as before and the location
of the ship track. If our ship passes directly through the center of a square, it will have
a relatively high amount of effort assigned, but if a square falls barely within the detection
distance of the trackline it will have a much smaller amount of effort assigned. Each detection
is then naturally assigned to the square in which it is located. The size of the squares can
be adjusted to be as small as desired, and thus can accomodate any level of granularity in
the environmental variables to be used in the study.

## Package Functionality

The aim of this package is to take in a dataset with GPS coordinates of a ship track and a
set of localized detections and create an output that can be used as input for a spatial
density model. This means that for each detection we need to know the appropriate amount
of effort to associate with that detection, and the location to associate with that detection
so that any environmental variables can later be assigned with other functions (such as the
`matchEnvData` function from the packge `PAMmisc`). 

The first step in this process is to create the grid. We only need to know the total size
of the grid to create, and how large each individual square should be. Once the grid is created,
we need to use the path of the ship and the detection function to estimate how much effort
should be placed in each grid. This is done (in rough terms) by scaling the area surveyed
by the probability of the detection function. Since detection probability typically 
decreases with distance, this means grid squares further from the track line will have
less effort assigned to them. This step can be very time consuming as it involves many
thousands of calculations across the entire grid, so there is also an intermediate step
to speed this up. This step involves "connecting" the grid so that each square knows which
squares it is adjacent to, and it vastly speeds up the processing time required for the
effort calculations. Once this effort calculation is done, then the last step is to
see how many detections fall within each grid cell, then our work is done. 

## Data Formatting Requirements

The ship track data must have columns `Longitude`, `Latitude`, `UTC` - the time in POSIXct
format with time zone UTC, and `effort` - a logical flag indicating which portions of the
ship track line are considered "On Effort". Off effort locations (marked with `FALSE`) will
not be used for determining effort. 

The detection data only needs columns `Longitude` and `Latitude` for the localized detection 
location, but if you want to create a detection function using the `Distance` package then it 
also needs a column `distance`, the perpendicular distance to the trackline in meters. Alternatively
you can provide your own detection function that takes in a distance as input and outputs a detection
probability.

## Package Useage

This package is designed to be extremely user friendly. Once data is in the appropriate format, all
of the gridding process can be accomplished with a single function call. First we read in our data,
and examine it to make sure we have the necessary columns.

```{r echo=TRUE, cache=TRUE}
# First load the functions required
source('./devel/gridFunctions.R')
gpsData <- readRDS('ExampleGPS.Rdata')
str(gpsData)
detections <- readRDS('ExampleDetections.RData')
str(detections)
```

Next we will use the `Distance` package to create a detection function from our detection data.
Here we'll fit a half-normal detection function and plot the result.

```{r, echo=TRUE, cache=TRUE}
library(Distance)
dsModel <- ds(detections, key='hn')
plot(dsModel)
```

Now we are ready to calculate the gridded effort. This just requires two more decisions. First we need to
decide what size the individual grid squares should be. This is set with the `pixel` argument, and
the units of this are in decimal degrees. We'll create a grid of squares with side length of 0.01 degrees
for our example, which is approximately 1km x 1km squares. For most applicatons this would be too fine of
a resolution, usually you would select a size that fits the resolution of any environmental variables you would
later want to include. So if you wanted to include SST from a dataset that has spatial resolution of 5km, 
it would make sense to select a pixel size of approximately 5km x 5km. If no value is input for `pixel`, then
the user will be prompted to select from a few pre-determined options. Second we need to set a truncation distance,
just like with normal line transect distance sampling. For most applications in acoustics this distance will
be higher than with visual surveys. This is a synthetic dataset created just for this example, and the detections
are limited to 10km away, so we will use `10e3` for our truncation distance. 

This function can take quite a bit of time to run, especially for longer tracklines and smaller pixel sizes, so 
it includes some progress bars to reassure you that your R session hasn't frozen and progress is being made.
The progress bar has been omitted from this document because it looks awful in RMarkdown, but they'll be there
when you run the function.

```{r echo=TRUE, results='hide', cache=TRUE}
gridDataset <- doAllGrid(gps = gpsData,
                      dets = detections,
                      trunc_m = 10e3, #METERS
                      dsmodel=dsModel,
                      pixel = .01,
)
``` 

This creates an output list with 5 items:

```{r echo=TRUE}
names(gridDataset)
```

`gps` is the same `gpsData` object we used as input

```{r echo=TRUE}
str(gridDataset$gps)
```

`grid` is the grid we created, this can be fed back in to future calls to `doAllGrid` to save
some time as long as the location and pixel sizes are the same (maybe if you wanted to try
 a different detection function, or were running multiple models for different species on the
 same trackline). It is a SpatialFeatures POLYGON object (package `sf`) that has the additional
 class of `"connectedGrid"` specific to this package.

```{r echo=TRUE}
class(gridDataset$grid)
```

`effort` is a list that stores the amount of effort (roughly in units of area meters^2) in each grid
cell. It also keeps track of some other stuff that is mostly used for debugging. The area is indexed 
by grid cell number, so the first area value corresponds to the first grid cell. 

```{r echo=TRUE}
names(gridDataset$effort)
```

`detections` is the same as the `detections` dataframe we supplied as input, but with a 
few new columns added. `gridIndex` is the grid number the detection is located in,
`effortArea` is the amount of effort in that grid, `actualArea` is the actual area
of the grid cell. Note that it is possible for `effortArea` to be greater than
`actualArea` if a survey has passed nearby a location multiple times. `gridLongitude`
and `gridLatitude` are the coordinates of the centroid of the grid cell a detection
is located in. 

```{r echo=TRUE}
str(gridDataset$detections)
```

`dataset` is a dataframe formatted in the way that a modeling dataset likely would be. It has
one row for every grid cell. `gridIx` is the grid cell index number, `nDetections` is the total
number of detections in that grid, `Longitude` and `Latitude` are the coordinates of the 
centroid of each grid cell (similar to `gridLongitude` and `gridLatitude` from `detections`),
and `effort` is the amount of effort in that grid cell. This dataframe would need to be further
augmented by environmental variables in order to be useful.

The package also contains a plotting function to show the result of the gridded effort calculations.

```{r echo=TRUE, cache=TRUE}
plotGridResult(gridDataset)
```

This plot shows our synthetic survey trackline in blue. Note that the middle section was marked as 
off effort for this test dataset to mimic how ship based surveys transit between two tracklines, so
only the two parallel tracks are contributing to effort. Grid cells are shaded according to the amount
of effort, white being no effort, and darkest grays corresponding to an amount of effort equal to
or greater than the area of the grid cell. We see that areas near the track line are darker gray, and then 
as we get further it fades to white. The areas in between the two tracklines get effort from both
tracklines and are darker than areas a similar distance away that received effort from only one track.
The red dots signify the locations of our detections.